---
title: "Plot sizes with `ggsave()`"
author: "Stefano Coretta"
date: "4/13/2021"
output: 
  html_document: 
    highlight: zenburn
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
```

The default font size (`base_size`) is 11 pt (see `?theme_classic`).

This applies to `element_text()` and `geom_text/label()`.

The font size in `element_text()` is in pt while in `geom_text/label()` has to be specified as `11 / .pt`.

```{r font-size}
# A base plot.
p <- ggplot(mtcars, aes(wt, mpg, label = rownames(mtcars))) +
  labs(
    x = "Weight", y = "Mileage", title = "The title", subtitle = "The sub", caption = "The caption."
  )

# Add geom_text().
p1 <- p  + geom_text()
ggsave("plot-sizes/p1.png", p1, width = 7, height = 5)

# Specify font size.
p2 <- p + geom_text(size = 11 / .pt) + theme(text = element_text(size = 11))
ggsave("plot-sizes/p2.png", p2, width = 7, height = 5)

# p1 and p2 are the same
```

```{r}
p3 <- p + geom_text(size = 9 / .pt) + theme(text = element_text(size = 16))
ggsave("plot-sizes/p3.png", p3, width = 7, height = 5)
```


The default dpi is 300 (note that it's 72 with knitr).
Changing the dpi changes the resolution if width/height is the same.

The width and height, if not specified, are those returned by `dev.size()` which is the current size of the `Plots` tab in RStudio.

```{r set-dpi}
ggsave("plot-sizes/p1-300.png", p1, width = 7, height = 5)
ggsave("plot-sizes/p1-72.png", p1, width = 7, height = 5, dpi = 72)
```

